---
title: "Project: Dealing with Polygon Data"
subtitle: <h4 style="font-style:normal">GEOG-364 - Spatial Analysis</h4>
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    theme: flatly
---


<style>
p.comment {
background-color: #DBDBDB;
padding: 10px;
border: 1px solid black;
margin-left: 0px;
border-radius: 5px;
font-style: normal;
}

h1.title {
  font-weight: bold;
  font-family: Arial;  
}

h2.title {
  font-family: Arial;  
}

</style>


<style type="text/css">
#TOC {
  font-size: 11px;
  font-family: Arial;
}
</style>


```{r, results=FALSE,warning=FALSE,message=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message = FALSE)
```


```{r,message=FALSE,warning=FALSE,results=FALSE,include=FALSE}
library(sp)
library(sf)
library(elevatr)
library(USAboundaries)
library(raster)
library(tidycensus)
library(rnaturalearth)
library(tmap)
library(exactextractr)
library(units)
library(tigris)
library(rmapshaper)
library(FedData)
```


```{r setup, include=FALSE,message=FALSE,warning=FALSE,results=FALSE}

# SVI shapefile data
SVI_county.sf <- st_read("SVI2018_IOWA_county.shp")
SVI_county.sf <- st_transform(SVI_county.sf,32614)

# Census data
ACS_county.sf <- get_acs(geography = "county", 
                         year = 2019,
                         variables = c(housevalue       = "B25075_001",  # house value
                                       total_pop        = "B05012_001",  # total population
                                       total_house      = "B25001_001",  # total housing units  
                                       med.income       = "B19013_001"), # median income  
                         state = c("IA"),
                         survey = "acs5",geometry=TRUE,
                         output = "wide")

ACS_county.sf  <- st_transform(ACS_county.sf,4326)

ACS_county.sf$County.areaE <- st_area(ACS_county.sf)
ACS_county.sf$County.areaE <- set_units(ACS_county.sf$County.areaE,"km^2")
ACS_county.sf$PopulationDensityE  <- ACS_county.sf$total_popE /  ACS_county.sf$County.areaE
ACS_county.sf$Log_PopulationDensityE <- log(ACS_county.sf$PopulationDensityE)

border.county.IA.sf <- us_counties(states="Iowa")
elevation.raster_IA <- elevatr::get_elev_raster(border.county.IA.sf, z = 5,clip="locations")
IA_Landcover <- raster('IA_Landcover.tif')


```

## Cropping data/zooming in

### Required libraries

```{r,message=FALSE,warning=FALSE,results=FALSE,eval=FALSE}
library(sp)
library(sf)
library(elevatr)
library(USAboundaries)
library(raster)
library(tidycensus)
library(rnaturalearth)
library(tmap)
library(exactextractr)
library(units)
library(tigris)
library(rmapshaper)
library(FedData)
```

### Filtering point data sf

I am basing my examples on Iowan SVI data, SVI_county.sf

```{r}
tmap_mode("plot")

tm_shape(SVI_county.sf)+
  tm_polygons("E_MOBILE",title="Estimate Living in a Mobile home/county",palette="RdPu")+
    tm_layout(legend.outside=TRUE)


```

Make sure your data is in sf format!   Now there are a few options. You could filter to one of the columns and crop that way. For my data I have these columns

```{r}
names(SVI_county.sf)
```

So for example

```{r}
#or
SVI_county.UIOWA.sf <- dplyr::filter(SVI_county.sf, COUNTY %in% c("Linn","Johnson"))

map1 <- tm_shape(SVI_county.sf)+
  tm_polygons("E_MOBILE",title="Estimate Living in a Mobile home/county",palette="Reds")+
    tm_layout(legend.outside=TRUE)


map2 <- qtm(st_geometry(SVI_county.sf))+
           tm_shape(SVI_county.UIOWA.sf)+
             tm_polygons("E_MOBILE",title="Estimate Living in a Mobile home/county",palette="Reds")+
           tm_layout(legend.outside=TRUE)


map3 <- tm_shape(SVI_county.UIOWA.sf)+
  tm_polygons("E_MOBILE",title="Estimate Living in a Mobile home/county",palette="Reds")+
    tm_layout(legend.outside=TRUE)

tmap_arrange(map1,map2,map3)
rm(map1);rm(map2); rm(map3)

                                                  
```

### Cropping to a lon/lat box

Ocean folks, this one is for you!

```{r}

# Make sure your sf data is in lon/lat
SVI_county.sf  <-  st_transform(SVI_county.sf,4326)

# Turn it into sp data
SVI_county.sp <- as(SVI_county.sf,"Spatial")

# My new region from https://www.geoplaner.com/
Crop.Region <- as(extent(-94,-92,41,42), "SpatialPolygons")

# Subset the polygons to my new region
SVI_county.BOX.sp <- crop(SVI_county.sp, Crop.Region, byid=TRUE)

# and convert back to sf and your other map projection
SVI_county.BOX.sf <- st_as_sf(SVI_county.BOX.sp)

SVI_county.sf  <-  st_transform(SVI_county.sf,3744)
SVI_county.BOX.sf  <-  st_transform(SVI_county.BOX.sf,3744)


# and plot -

map1 <- tm_shape(SVI_county.sf)+
             tm_polygons("E_TOTPOP",title="Estimate of the total population",
                         palette="Blues",style="log10_pretty")+
        tm_layout(legend.outside=TRUE)

map2 <- tm_shape(SVI_county.sf)+tm_borders()+
        tm_shape(SVI_county.BOX.sf)+
             tm_polygons("E_TOTPOP",title="Estimate of the total population",
                         palette="Blues",style="log10_pretty")+
        tm_layout(legend.outside=TRUE)

map3 <- tm_shape(SVI_county.BOX.sf)+
             tm_polygons("E_TOTPOP",title="Estimate of the total population",
                         palette="Blues",style="log10_pretty")+
        tm_layout(legend.outside=TRUE)

tmap_arrange(map2,map1,map3)
rm(map1);rm(map2);rm(map3)

```

### Cropping to a state/shapefile

Sometimes we want to crop to a state/shapefile.  For example imagine we want to crop to a city metropoliton area (see Lab 7)

```{r}

# This could be any sf boundary you want to crop to.
# it doesn't have to be a city - here i look for Des Moines
pl <- places(state = "IA", cb = TRUE, year=2017)
DesMoines.borders.sf <- dplyr::filter(pl, NAME == "Des Moines")

# make sure they are the same map projection
SVI_county.sf  <-  st_transform(SVI_county.sf,4326)
DesMoines.borders.sf  <-  st_transform(DesMoines.borders.sf,4326)

# and crop and save to a new thing
DesMoines.SVI.sf <- ms_clip(target = SVI_county.sf, clip = DesMoines.borders.sf, remove_slivers = TRUE)

# change the map projection back
SVI_county.sf  <-  st_transform(SVI_county.sf,3744)
DesMoines.borders.sf  <-  st_transform(DesMoines.borders.sf,3744)
DesMoines.SVI.sf  <-  st_transform(DesMoines.SVI.sf,3744)


# and plot

map1 <- qtm(st_geometry(border.county.IA.sf),fill="white")+
        tm_shape(DesMoines.SVI.sf)+
             tm_polygons("E_AGE65",title="Estimate of over 65s",
                         palette="Blues",style="log10_pretty")+
        tm_shape(DesMoines.borders.sf)+tm_borders()


map2 <- tm_shape(DesMoines.SVI.sf)+
             tm_polygons("E_AGE65",title="Estimate of over 65s",
                         palette="Blues",style="log10_pretty")+
        tm_shape(DesMoines.borders.sf)+tm_borders()

tmap_arrange(map1,map2)
rm(map1);rm(map2)
```


## Adding in secondary data

### Required libraries

```{r,message=FALSE,warning=FALSE,results=FALSE,eval=FALSE}
library(sp)
library(sf)
library(elevatr)
library(USAboundaries)
library(raster)
library(tidycensus)
library(rnaturalearth)
library(tmap)
library(exactextractr)
library(units)
library(tigris)
library(rmapshaper)
library(FedData)
```


### Averaging point values over polygons

There is a great tutorial here: 

This is very easy using the `st_join()` command from the `sf` package. To add my census data (or any sf polygon data) to my Iowa SVI data

```{r}
#--------------------------------------------------------------
# Transform all to lon/lat crs=4326 - just in case
#--------------------------------------------------------------
ACS_county.sf        <-  st_transform(ACS_county.sf,4326)
SVI_county.sf  <-  st_transform(SVI_county.sf,4326)

#--------------------------------------------------------------
# MAKE SURE THE POINT DATA GOES FIRST
# I'm calling it something new so I dont accidentally overwrite
#--------------------------------------------------------------
SVI_county.sf.all <- st_join(SVI_county.sf,ACS_county.sf)

#--------------------------------------------------------------
# Transform back
#--------------------------------------------------------------
ACS_county.sf        <-  st_transform(ACS_county.sf,3744)
SVI_county.sf  <-  st_transform(SVI_county.sf,3744)
SVI_county.sf.all <- st_transform(SVI_county.sf.all,3744)

#--------------------------------------------------------------
# and plot
#--------------------------------------------------------------
tmap_mode("plot")

map1 <- tm_shape(SVI_county.sf.all)+
             tm_polygons("E_AGE65",title="Estimate of over 65s",
                         palette="Blues",style="log10_pretty")


map2 <- tm_shape(SVI_county.sf.all)+
             tm_polygons("housevalueE",title="Estimate of house value",
                         palette="Spectral",style="log10_pretty")

tmap_arrange(map1,map2,ncol=2)

# clean up your workspace, optional but good practice unless you later need this map
rm(map2); rm(map1)
```


### Extracting polygon values from rasters/fields

A raster is a field like elevation or land cover.

```{r,include=FALSE}

tmap_mode("plot")

map_elev     <- qtm(elevation.raster_IA,title="Elevation")+
        tm_shape(IA_farmers_census.sf)+tm_dots(size=.1)+
        tm_shape(border.county.IA.sf)+tm_polygons(alpha=0)+
        tm_layout(legend.outside = TRUE,legend.outside.position	 = "bottom")

map_land <- qtm(IA_Landcover,title="Land Cover")+
        tm_shape(IA_farmers_census.sf)+tm_dots(size=.1)+
        tm_shape(border.county.IA.sf)+tm_polygons(alpha=0)+
        tm_layout(legend.outside = TRUE,legend.outside.position	 = "right")

tmap_arrange(map_elev,map_land,ncol=2)

# clean up your workspace, optional but good practice unless you later need this map
rm(map_elev); rm(map_land)
```

Here the elevation data is numeric whilst the land cover is categorical.

#### Extracting polygon values from numeric rasters 

This is very easy using the `exact_extract()` command from the `exactextractr` package. To extract my elevation data 

For point data, we choose simply the extract function.  Remember to put results=FALSE

```{r, results=FALSE,warning=FALSE}
#--------------------------------------------------------------
# Turn everything to the same map projection
#--------------------------------------------------------------
SVI_county.sf  <-  st_transform(SVI_county.sf,crs(elevation.raster_IA))

#--------------------------------------------------------------
# add to polygon data
#--------------------------------------------------------------
SVI_county.sf$ElevationMean <- exact_extract(elevation.raster_IA,SVI_county.sf,"mean")
SVI_county.sf$ElevationMax <- exact_extract(elevation.raster_IA,SVI_county.sf,"max")
SVI_county.sf$ElevationMin <- exact_extract(elevation.raster_IA,SVI_county.sf,"min")

#--------------------------------------------------------------
# Turn everything to the final map projection
#--------------------------------------------------------------
SVI_county.sf <-  st_transform(SVI_county.sf,3744)
```


```{r, warning=FALSE}
#--------------------------------------------------------------
# And plot
# see here for more colours: https://www.r-graph-gallery.com/38-rcolorbrewers-palettes.html
#--------------------------------------------------------------
tmap_mode("plot")

map_elev     <- tm_shape(border.county.IA.sf)+
                    tm_polygons(alpha=0)+
                tm_shape(elevation.raster_IA)+
                     tm_raster(title="Elevation",
                               palette="Spectral",style="cont",breaks=seq(100,600,100))+
               tm_shape(border.county.IA.sf)+
                    tm_polygons(alpha=0)+
                tm_layout(legend.outside = TRUE,legend.outside.position	 = "bottom")
                
map2 <- tm_shape(SVI_county.sf)+
             tm_polygons("ElevationMean",title="Mean Elevation in each county",
                         palette="Spectral",style="cont",breaks=seq(100,600,100))+
        tm_layout(legend.outside=TRUE,legend.outside.position	 = "bottom")


map3 <- tm_shape(SVI_county.sf)+
             tm_polygons("ElevationMax",title="Max Elevation in each county",
                         palette="Spectral",style="cont",breaks=seq(100,600,100))+
        tm_layout(legend.outside=TRUE,legend.outside.position	 = "bottom")

map4 <- tm_shape(SVI_county.sf)+
             tm_polygons("ElevationMin",title="Min Elevation in each county",
                         palette="Spectral",style="cont",breaks=seq(100,600,100))+
        tm_layout(legend.outside=TRUE,legend.outside.position	 = "bottom")

tmap_arrange(map_elev,map2,map3,map4,ncol=2)
rm(map_elev); rm(map2);rm(map3);rm(map4)
```

#### Extracting point values from categorical rasters

This is exactly the same but we use the "majority" function

```{r, results=FALSE,warning=FALSE}
#--------------------------------------------------------------
# Turn everything to the same map projection
#--------------------------------------------------------------
SVI_county.sf         <-  st_transform(SVI_county.sf,crs(IA_Landcover))

#--------------------------------------------------------------
# add to polygon data
#--------------------------------------------------------------
SVI_county.sf$LandcoverNumber <- exact_extract(IA_Landcover,SVI_county.sf,"majority")

#--------------------------------------------------------------
# for the landcover data specifically, this might be better
#--------------------------------------------------------------
key <- nlcd_colors()[,c("ID","Class")]
names(key) <- c("LandcoverNumber","LandcoverClass")

SVI_county.sf <- merge(SVI_county.sf ,key,by="LandcoverNumber",all.x=TRUE,all.y=FALSE)

SVI_county.sf$LandcoverNumber <- as.factor(SVI_county.sf$LandcoverNumber )
SVI_county.sf$LandcoverClass <- as.factor(SVI_county.sf$LandcoverClass )

#--------------------------------------------------------------
# Turn everything to the final map projection
#--------------------------------------------------------------
SVI_county.sf <-  st_transform(SVI_county.sf,3744)

#--------------------------------------------------------------
# And plot
# see here for more colours: https://www.r-graph-gallery.com/38-rcolorbrewers-palettes.html
#--------------------------------------------------------------
tmap_mode("plot")

map_lc1     <- tm_shape(border.county.IA.sf)+
                    tm_polygons(alpha=0)+
                tm_shape(IA_Landcover)+
                     tm_raster(title="Landcover")+
               tm_shape(border.county.IA.sf)+
                    tm_polygons(alpha=0)+
                tm_layout(legend.outside = TRUE,legend.outside.position	 = "bottom")
                
map_lc2 <- tm_shape(SVI_county.sf)+
             tm_polygons("LandcoverNumber",title="Majority Land cover code")+
        tm_layout(legend.outside=TRUE,legend.outside.position	 = "bottom")

                
map_lc3 <- tm_shape(SVI_county.sf)+
             tm_polygons("LandcoverClass",title="Majority Land cover class")+
        tm_layout(legend.outside=TRUE,legend.outside.position	 = "bottom")



tmap_arrange(map_lc1,map_lc2,map_lc3,ncol=3)
rm(map_lc1); rm(map_lc1); rm(map_lc3)
```



