---
title: "Project 1"
subtitle: <h4 style="font-style:normal">GEOG-364 - Spatial Analysis</h4>
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    theme: flatly
---

<style>
p.comment {
background-color: #DBDBDB;
padding: 10px;
border: 1px solid black;
margin-left: 0px;
border-radius: 5px;
font-style: normal;
}

h1.title {
  font-weight: bold;
  font-family: Arial;  
}

h2.title {
  font-family: Arial;  
}

</style>


<style type="text/css">
#TOC {
  font-size: 12px;
  font-family: Arial;
}
</style>

\
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message = FALSE)
```


```{r,include=FALSE,echo=FALSE}
# invisible data read
library(tidyverse)
library(sp)
library(sf)
library(readxl)
library(skimr)
library(tmap)
library(USAboundaries)
library(viridis)
library(kableExtra)

frost <- readxl::read_excel("pg_364Data_1frostday.xlsx")
```

## Final project 1. Topic choice

<br>

in lieu (instead of) a final exam, you will be conducting a long-term project for this class, where you will independently use the skills you have learned in this class and formally write up the results in Markdown.  

There are three separate stages to this
 1. Topic choice
 2. Exploratory statistics
 3. Final analysis and write up.

The first deadline defines and explores your topic idea. Additional deadlines build on this initial submission. 

This is due by midnight the night before your next lab on Canvas. 

See [**this page**]https://psu.instructure.com/courses/2120046/assignments/13274854) on canvas for assignment guidelines.



***

Website created and maintained by [Helen Greatrex](https://www.geog.psu.edu/directory/helen-greatrex). Website template by [Noli Brazil](https://nbrazil.faculty.ucdavis.edu/)