---
title: "Tutorial 8: Summary Statistics"
subtitle: <h5 style="font-style:normal">GEOG-364 - Spatial Analysis</h4>
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    theme: flatly
---


<style>
p.comment {
background-color: #DBDBDB;
padding: 10px;
border: 1px solid black;
margin-left: 0px;
border-radius: 5px;
font-style: normal;
}

h1.title {
  font-weight: bold;
  font-family: Arial;  
}

h2.title {
  font-family: Arial;  
}

</style>


<style type="text/css">
#TOC {
  font-size: 12px;
  font-family: Arial;
}
</style>

\
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message = FALSE)
```

## Tutorial 8: Summary Statistics

This tutorial is all about exploratory data analysis

 - [**Tutorial 8A: What is a data.frame?**](#Tut8_whatisit)
  <br>
      a. [Understanding the data itself](#Tut8a1_basics)
      b. [Summary Statistics](#Tut8a2_summary)
<br>

<div style="margin-bottom:25px;">
</div>  
### Data.frame introduction {#Tut8a_whatisit}

Most of the data we will look at is in "data.frame" format.  This is a table, just like an excel spreadsheet, with one row for each observation and one column for each variable. Each column has a column name.

In this tutorial, I will focus on in-built R datasets. 

Let's choose one now. I'm going to work with the pirates dataset from the `yarrr` package. We can choose the data here.

```{r, echo=FALSE,message=FALSE,warning=FALSE, include=FALSE}
library(yarrr)
library(tidyverse)
piratedataset <- pirates
```

```{r, eval=FALSE, echo=TRUE}
library(yarrr)
library(tidyverse)
?pirates
piratedataset <- pirates
```


### Looking at the data itself {#Tut8a1_basics}


To have a look at the data there are many options. You can:

 - click on its name in the environment tab
 - Type its name into the console or into a code chunk (e.g. for our table, type `piratedataset` into the console or a code chunk)
 - Run the command `View(variable_name)` (View is a command from the tidyverse package).<br> This will open the data in a new tab.
 - Run the command `head(variable_name)` to see the first 6 lines or so (good for quick checks)
 - Run the command `glimpse(variable_name)` to get a nice summary.
 - Run the command `names(variable_name)` to get the column names.
 - 
 
<br>
 
For example
 
```{r}
# Note, there are more columns to the right, use the arrow to see
head(piratedataset)
```
To see what the column names are, you can use the `names(dataset)` command

```{r}
names(piratedataset)
```

Or the glimpse command:

```{r}
glimpse(piratedataset)
```
To see how many columns and rows there are, you can use the `nrow()` and `ncol()` commands

```{r}
nrow(piratedataset)
ncol(piratedataset)
```

### Summary statistics {#Tut8a2_summary}

To look at the summaries there are a load of options. Choose your favourites:

 - `summary(dataset)`
 - `skim(dataset)` in the skimr package
 - `summarize(dataset)` in the papeR package. This looks pretty powerful, I'm just learning it

Or look at the summary

```{r}
summary(piratedataset) 
```

In the skimr library there is the skim command

```{r}
library(skimr)
skim(piratedataset) 
```





<br>
<br>


***

Website created and maintained by [Helen Greatrex](https://www.geog.psu.edu/directory/helen-greatrex). Website template by [Noli Brazil](https://nbrazil.faculty.ucdavis.edu/)